---
title: "Group 18 - R Bootcamp"
author: "Noel Rinke, Yanik Baumann"
date: \today
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    includes:
      in_header: "wrap-code.tex"
bibliography: Sources.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Noel Rinke and Yanik Baumann are two students at the Lucerne University of Applied Sciences and Arts studying for a Master's degree in Applied Information and Data Science. Our assignment represents a complete data analysis where we use a wide variety of R functionalities. For this analysis, we use data from Citibike New York. 

## Background Information on New York's Bike-Sharing Scheme (work-in-progress)

As the New York Times [-@Hu2021] reported "*traffic has come roaring back in New York City after largely disappearing at the height of the corona virus pandemic*". While cars contribute to traffic congestion, greenhouse gas emissions and global warming, bicycles offer an environmentally friendlier, healthier, more economical and less space-consuming transportation alternative. Due to these benefits, bike-sharing schemes are increasingly promoted and expanded in urban areas such as the bike-sharing service Citibike in New York City provided by the company Lyft. 

## Aim of the Analysis

blabla


## Datasets used

The subsequent analysis required two different datasets: ride-sharing data and weather data.

The data with records on the bicycle rides was obtained from citibike, a bike-sharing service operating in New York City. To represent each season, four different data sets were downloaded for January, April, July, and December, and an equal sample size was drawn from each month in the data preparation section. In addition, it is important to note that the Covid 19 pandemic was thought to be a potential confounding factor, so we specifically chose to use data from 2019, which predated the pandemic.

The second dataset used stems from the National Oceanic and Atmospheric Administration which is part of the U.S. Department of Commerce. The dataset includes weather data measured at the Central Park in New York which corresponds to the weather station with the number GHCND:USW00094728.


# Installation of Packages

```{r packages, echo=TRUE, warning=FALSE}
library(dplyr)

```

# Data Preparation

bla


## Data Import

```{r data import, echo=TRUE, eval=FALSE, tidy=TRUE}
citibike_jan19 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\201901-citibike-tripdata.csv'))

citibike_apr19 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\201904-citibike-tripdata.csv'))

citibike_jul19 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\201907-citibike-tripdata.csv'))

citibike_oct19 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\201910-citibike-tripdata.csv'))

```

```{r data import (weather data)}
nyc_weather_2019 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\New York Weather 2019.csv'))

```


## Taking Samples

```{r sample, echo=TRUE, eval=FALSE,}
citibike_jan19_sample <- sample_n(citibike_jan19, size=1000)
citibike_apr19_sample <- sample_n(citibike_apr19, size=1000)
citibike_jul19_sample <- sample_n(citibike_jul19, size=1000)
citibike_oct19_sample <- sample_n(citibike_oct19, size=1000)

```


## Merging the Datasets

```{r dataset merge, echo=TRUE, eval=FALSE,}
citibike_merged_2019 <- rbind(citibike_jan19_sample, citibike_apr19_sample, citibike_jul19_sample, citibike_oct19_sample)

```

```{r create csv with all created samples, echo=TRUE, eval=FALSE,}
write.csv(citibike_merged_2019, 'citibike_merged_2019.csv')

```


# Analysis



## Data Inspection


## Frequencies


## Density Map: Identifiying the most and least popular Districts


## Influence of Meteorological Factors


### Influence of Temperature


### Influence of Precipitation


### Influence of Snowfall and Snow Depth



# Results



# Sources


