---
title: "Group 18 - R Bootcamp"
author: "Noel Rinke, Yanik Baumann"
date: \today
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    includes:
      in_header: "wrap-code.tex"                  # Wraps code (when not fitting on one line)
    latex_engine: xelatex                         # Solves encoding issue
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm" # Customize margins
biblio-style: apa
bibliography: Sources.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Noel Rinke and Yanik Baumann are two students at the Lucerne University of Applied Sciences and Arts studying for a Master's degree in Applied Information and Data Science. Our assignment represents a complete data analysis where we use a wide variety of R functionalities. For this analysis, we use data from Citibike New York. 

## Background Information on New York's Bike-Sharing Scheme (work-in-progress)

The bike sharing scheme Citi Bike began operating in New York City in 2013 [@LyftInc22]. With 19000 bikes and more than 1000 stations (as of 2022), Citi Bike is the largest bike-sharing service in the United States, making it an integral part of New York City's transport network [@LyftInc22]. The bike-sharing scheme consists of a fleet of specifically engineered, long-lasting bicycles that are connected to a network of stations which span with Manhattan, Brooklyn, Queens and the Bronx four of the five boroughs of New York City as well as Jersey City and Hoboken. Citi Bike provides an app which allows its customers to unlock bikes, see real-time station information, and personalized data such as kilometers traveled and calories burnt.

As the New York Times [-@Hu2021] reported "*traffic has come roaring back in New York City after largely disappearing at the height of the corona virus pandemic*". While cars contribute to traffic congestion, greenhouse gas emissions and global warming, bicycles offer an environmentally friendlier, healthier, more economical and less space-consuming transportation alternative.


## Aim of the Analysis

blabla


## Datasets used

The subsequent analysis required two different datasets: Citi Bike ride-sharing data and weather data.

As mentioned above, the data with records on the bicycle rides was obtained from Citi Bike which provides all datasets since the service began operating in 2013. To represent each season, four different data sets were downloaded for January, April, July, and December, and an equal sample size was drawn from each month in the data preparation section. In addition, it is important to note that the Covid 19 pandemic was thought to be a potential confounding factor, so we specifically chose to use data from 2019, which predated the pandemic.

The second dataset used stems from the National Oceanic and Atmospheric Administration which is part of the U.S. Department of Commerce. The dataset includes weather data measured at the weather station with the number GHCND:USW00094728 which corresponds to New York City's Central Park. Important to note is that the weather data provided does not include measurements per hour, but per day.


# Installation of Packages

```{r, echo=TRUE, warning=FALSE, message=FALSE}
require(maps)
require(leaflet)
require(skimr)
require(st)
require(sf)
require(leaflet.extras)
require(ggeasy)
require(tidyr)
require(ggplot2)
require(ggthemes)
require(dplyr)
require(scales)
require(knitr)
require(lubridate)
require(ggmap)
require(rgdal)
require(httr)
require(broom)
```


# Data Preparation

This chapter describes how to import and prepare the data sets.


## Data Import

```{r}
citibike_jan19 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\201901-citibike-tripdata.csv'))

citibike_apr19 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\201904-citibike-tripdata.csv'))

citibike_jul19 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\201907-citibike-tripdata.csv'))

citibike_oct19 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\201910-citibike-tripdata.csv'))

```

```{r}
nyc_weather_2019 <- read.csv(file = file.path('C:\\Users\\noelr\\switchdrive',
'\\SharedOstSchweiz\\R Bootcamp\\Data\\New York Weather 2019.csv'))

```


## Taking Samples

```{r}
citibike_jan19_sample <- sample_n(citibike_jan19, size=1000)
citibike_apr19_sample <- sample_n(citibike_apr19, size=1000)
citibike_jul19_sample <- sample_n(citibike_jul19, size=1000)
citibike_oct19_sample <- sample_n(citibike_oct19, size=1000)

```

## Adding Season to the data sets

```{r}
citibike_jan19_sample <- cbind(citibike_jan19_sample, season='winter')
citibike_apr19_sample <- cbind(citibike_apr19_sample, season='spring')
citibike_jul19_sample <- cbind(citibike_jul19_sample, season='summer')
citibike_oct19_sample <- cbind(citibike_oct19_sample, season='autumn')

```

## Merging the Datasets from Citibike

```{r}
citibike_merged_2019 <- rbind(citibike_jan19_sample, citibike_apr19_sample, citibike_jul19_sample, citibike_oct19_sample)

```

## Adding Date-column

```{r}
citibike_merged_2019$DATE <- substr(citibike_merged_2019$starttime, 1, 10)

```

## Cleaning the weather data set

```{r}
nyc_weather_cleaned <- subset(nyc_weather_2019, select = -c(TSUN, TAVG, AWND))

```

## Join citibike_merged_2019 and the nyc_weather_cleaned

```{r}
citi_cb <- left_join(citibike_merged_2019, nyc_weather_cleaned, 
                                       by='DATE')
```

## Calculate average temperature
```{r}
citi_cb$meantemp <- (citi_cb$TMAX + citi_cb$TMIN) / 2
```

## Grouping rain intensity
```{r}
citi_cb$raingrouped <- ifelse(citi_cb$PRCP == 0, 'no rain',
                       ifelse(citi_cb$PRCP > 0   & citi_cb$PRCP <=2.0, 'weak',
                       ifelse(citi_cb$PRCP >2.0  & citi_cb$PRCP <=10.0, 'moderate',
                       ifelse(citi_cb$PRCP >10.0 & citi_cb$PRCP <=30.0, 'heavy', 'intense'))))
```

## Snow Yes/No
```{r}
citi_cb$Snow_yesno <- ifelse(citi_cb$SNOW > 0, 'yes', 'no')
```

## Getting the weekdays
```{r}
# Weekdays in english
Sys.setlocale("LC_TIME", "English")

# Define Weekday
citi_cb$weekday <- weekdays(as.Date(citi_cb$DATE))

# Setting correct factor order
citi_cb$weekday <- factor(citi_cb$weekday,levels = c("Monday","Tuesday","Wednesday",
                                                     "Thursday","Friday","Saturday","Sunday"))
```

## Getting Time and Daytimes
```{r}
citi_cb$onlytime <- as.POSIXct(substr(citi_cb$starttime, 12, 19), format="%H:%M:%S")

# Only Hours and Minutes
citi_cb$time     <- strftime(citi_cb$onlytime,format="%H:%M")

# Only Hour
citi_cb$hour     <- as.numeric(substr(citi_cb$time,1,2))

# Set Daytime
citi_cb$daytime  <- ifelse(citi_cb$onlytime <= as.POSIXct("06:00:00",format="%H:%M:%S"), 'night',
                    ifelse(citi_cb$onlytime > as.POSIXct("06:00:00",format="%H:%M:%S") & 
                    citi_cb$onlytime <= as.POSIXct("12:00:00",format="%H:%M:%S"), 'morning',
                    ifelse(citi_cb$onlytime > as.POSIXct("12:00:00",format="%H:%M:%S") & 
                    citi_cb$onlytime <= as.POSIXct("18:00:00",format="%H:%M:%S"), 'afternoon', 
                    ifelse(citi_cb$onlytime > as.POSIXct("18:00:00",format="%H:%M:%S") & 
                    citi_cb$onlytime <= as.POSIXct("24:00:00",format="%H:%M:%S"), 'evening', 'night'))))

# Setting correct factor order
citi_cb$daytime <- factor(citi_cb$daytime,levels = c("morning","afternoon","evening","night"))
```

## Setting trip duration groups
```{r}
citi_cb$tripduration_cat <- ifelse(citi_cb$tripduration <=300, '0 to 5 min',
                            ifelse(citi_cb$tripduration >300 & citi_cb$tripduration <=900, '5 to 15 min',
                            ifelse(citi_cb$tripduration >900 & citi_cb$tripduration <=1800, '15 to 30 min',
                            ifelse(citi_cb$tripduration >1800 &
                                   citi_cb$tripduration <=3600, '30 to 60 min','over 60 min'))))
```


## Outlier Detection and Elimination

Outliers can have a large impact on the statistics derived from the dataset, which is why they need to be handled or at least be aware of. The variable *tripduration* consists of the time (in seconds) a customer rents a bike from Citi Bike until the bike is returned to a station with an empty dock. Therefore, there can be both natural and non-natural outliers. While the non-natural outliers are due to measurement errors (e.g. system failures), natural outliers could be due to customers not returning the rented bike, customers not being able to return the rented bike until the next day, or similar.

```{r}
#Deactivate scientific notation
options(scipen = 999)

# Check if the variable trip duration has outliers
boxplot(citi_cb$tripduration, horizontal=TRUE)

```

The boxplot reveals that the variable *tripduration* contains two extreme outliers - one of about 64 hours and the other of about 278 hours. The fact that bicycles are rented for more than 24 hours can be considered as important information, but renting for several days is not in line with the purpose and design of the service. Although the cause of these outliers is not known, we decided to remove all outliers above 12 hours.

```{r}
# Remove outliers in the variable trip duration
citi_cb <- subset(citi_cb, tripduration < 43200)
boxplot(citi_cb$tripduration, horizontal=TRUE)

```


# Analysis



## Data Inspection
```{r}
# Alternative to summary to get an overview of the data frame
skim(citi_cb)

```

```{r}
# The 10 most popular stations
citi_cb %>% 
  group_by(start.station.name) %>% 
  summarise(count=n())

most_popular <- data.frame(citi_cb %>%
  group_by(start.station.name) %>% 
  summarize(count=n()) %>% 
  arrange(desc(count)))

most_popular_stations <- most_popular %>%
  left_join(citi_cb, by = "start.station.name") %>%
  select(start.station.name, count, start.station.longitude, start.station.latitude)

station_ranking_top10 <- head(distinct(most_popular_stations),10)
station_ranking_top10


```

```{r}
# The 10 least popular stations
least_popular <- data.frame(citi_cb %>%
                             group_by(start.station.name) %>% 
                             summarize(count=n()) %>% 
                             arrange(desc(count)))

least_popular_stations <- least_popular %>%
  left_join(citi_cb, by = "start.station.name") %>%
  select(start.station.name, count, start.station.longitude, start.station.latitude)

station_ranking_least10 <- head(distinct(least_popular_stations),10)
station_ranking_least10
```


## Frequencies


## Density Map: Identifiying the most and least popular Districts


## Influence of Meteorological Factors


### Influence of Temperature

```{r}
## How strong is the effect of the average daily temperature on the trip duration?
ggplot(citi_cb, aes(x=meantemp, y=tripduration)) +
  geom_point(color='#2980B9') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE, color='#2C3E50') +
  ggtitle(label="Effect of Average daily Temperature on Trip Duration") +
  ggeasy::easy_center_title() +
  xlab("Average daily temperature") +
  ylab("Trip duration (in seconds)")

# Spearman because there are two quantitative variables
cor(x=citi_cb$meantemp, y=citi_cb$tripduration, method="spearman")

```





### Influence of Precipitation

```{r}
# How strong is the effect of precipitation on the number of trips?

# Calculate for each category of precipitation how often it rained
rain_cat_freq <- citi_cb %>%
  group_by(raingrouped) %>%
  summarise(counts = n())

# Plot of the result
ggplot(rain_cat_freq, aes(x = raingrouped, y = counts)) +
  geom_bar(fill = "#0073C2FF", 
           stat = "identity") +
  geom_text(aes(label = counts), 
            vjust = -0.3) + # Label above bar
  xlab("Rain Category") +
  ylab("Number of Rides")
  
```

```{r, results="asis"}
## How strong is the effect of precipitation on the trip duration?
ggplot(citi_cb, aes(x=PRCP, y=tripduration)) +
  geom_point(color='#2980B9') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE, color='#2C3E50') +
  ggtitle(label="Effect of Precipitation on Trip Duration") +
  ggeasy::easy_center_title() +
  xlab("Precipitation (in mm)") +
  ylab("Trip duration (in seconds)")

cor_tripduration_precipitation <- round(cor(x=citi_cb$PRCP, 
                                            y=citi_cb$tripduration, 
                                            method="spearman"),2)
print(paste0("Interpretation: For every mm of precipitation, the time a customer rents a bike changes by ", cor_tripduration_precipitation, " seconds."))

```



### Influence of Snowfall and Snow Depth



# Results


# Lessons Learned

* To display maps in a pdf document or any other format except HTML, static maps need to be used. Leaflet - which was used initially - is a JavaScript library for building interactive maps for the web and hence not static. There is a workaround which is however not recommended to use.

# Sources


