---
title: "Group 18 - R Bootcamp"
author: "Noel Rinke, Yanik Baumann"
date: \today
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    includes:
      in_header: "wrap-code.tex"
bibliography: Sources.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

Noel Rinke and Yanik Baumann are two students at the Lucerne University of Applied Sciences and Arts studying for a Master's degree in Applied Information and Data Science. Our assignment represents a complete data analysis where we use a wide variety of R functionalities. For this analysis, we use data from Citibike New York. 

## Background Information on New York's Bike-Sharing Scheme (work-in-progress)

As the New York Times [-@Hu2021] reported "*traffic has come roaring back in New York City after largely disappearing at the height of the corona virus pandemic*". While cars contribute to traffic congestion, greenhouse gas emissions and global warming, bicycles offer an environmentally friendlier, healthier, more economical and less space-consuming transportation alternative. Due to these benefits, bike-sharing schemes are increasingly promoted and expanded in urban areas such as the bike-sharing service Citibike provided by the company Lyft. 


## Aim of the Analysis

blabla

## Datasets used

blabla

# Installation of Packages

```{r packages, echo=TRUE}

x <- 'This is a very very very very very very very very very very very very very very very very very very very very very very very very long text'


```

# Data Prepartion

bla

## Data Import

```{Data import, echo=TRUE, tidy=TRUE}

rides_jan <- read.csv(file = 'C:\\Users\\noelr\\switchdrive\\SharedOstSchweiz\\R Bootcamp\\Data\\201901-citibike-tripdata.csv')

```


## Merging the Datasets



# Data Analysis



## Data Inspection


## Frequencies


## Density Map: Identifiying the most and least popular Districts


## Influence of Meteorological Factors


### Influence of Temperature


### Influence of Precipitation


### Influence of Snowfall and Snow Depth


# Sources


