---
title: "Group 18 - R Bootcamp"
author: "Noel Rinke, Yanik Baumann"
date: \today
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Noel Rinke and Yanik Baumann are two students at the Lucerne University of Applied Sciences and Arts studying for a Master's degree in Applied Information and Data Science. Our assignment represents a complete data analysis where we use a wide variety of R functionalities. For this analysis, we use data from Citibike New York. 

## Background Information on Bike Sharing Schemes

blabla

## Aim of the Analysis

## Datasets used

# Installation of Packages

```{r packages, echo=FALSE}

```

# Data Prepartion



## Data Import



## Merging the Datasets



# Data Analysis



## Data Inspection


## Frequencies


## Density Map: Identifiying the most and least popular Districts


## Influence of Meteorological Factors


### Influence of average Temperature


### Influence of Precipitation


### Influence of Snowfall and Snow Depth





